---
title: "run_analysis.R Codebook"
author: "Michelle Piazza"
date: "February 4, 2018"
output: html_document
---
## Project Description
The purpose of this project is to demonstrate an ability to collect, work with, and clean a data set. The goal is to prepare tidy data that can be used for later analysis. 

##Study design and data processing
Data from the "Human Activity Recognition Using Smartphones Data Set" was used for this project.

###Collection of the raw data
Thirty subjects, ages 19-48 years, performed 6 activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. 561 features were recorded via the smartphone.

###Notes on the original (raw) data 
The data for the project can be found here:

https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

 
##Creating the tidy datafile

###Guide to create the tidy data file
1. create source directory if it does not already exist
```{r}
old.dir<-getwd()
if(!file.exists("./data")){dir.create("./data")}
```
2, download the zip file to the target directory
```{r}
datasetUrl<-"https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
download.file(datasetUrl,destfile = "./data/dataset.zip",mode="wb")
```
3. unzip the dataset
```{r}
unzip("./data/dataset.zip",exdir = "./data")
```
4. move to directory containing the dataset
```{r}
setwd("./data/UCI HAR Dataset")
```

###Cleaning of the data
1. Load the 8 text files using the read.table command
```{r}
activitylabel <- read.table("activity_labels.txt") ##there are 6 rows
featureslist <- read.table("features.txt")  ##there are 561 rows
testactivity<-read.table("./test/y_test.txt")  ##there are 2947 rows
trainingactivity<-read.table("./train/y_train.txt")  ##there are 7352 rows
testset<-read.table("./test/X_test.txt")  ##there are 2947 rows
trainingset<-read.table("./train/X_train.txt") ##there are 7352 rows
testsubject<-read.table("./test/subject_test.txt")    ##there are 2947 rows
trainingsubject<-read.table("./train/subject_train.txt") ##there are 7352 rows
```
2. combine the analagous data sets together via rbind()
   a. labels = activity
   b. sets = features
   c. subject = identifies the subjects
```{r}
activities <- rbind(testactivity,trainingactivity)  ##there are 10299 rows
features <- rbind(testset,trainingset)              ##there are 10299 rows
subjects <- rbind(testsubject,trainingsubject)      ##there are 10299 rows
```
3. convert the activity labels to readable factors
```{r}
activities$V1 <- factor(activities$V1, levels = activitylabel$V1, labels = activitylabel$V2)

```
4. give headers to the 3 data sets
```{r}
names(activities) <- "activity"
names(subjects) <- "subject"
names(features) <- featureslist$V2
```
5. find and subset the "mean" and "standard deviation" features
```{r}
featureswithmeanstddev <- as.character(featureslist$V2[grep("mean\\(\\)|std\\(\\)", featureslist$V2)])
featuressubset <- subset(features,select=featureswithmeanstddev)  ##length = 66, rows 10299

```
6. combine the subsetted data with the subjects and activites
```{r}
subjectbyactivity <- cbind(subjects, activities) ##10299 rows, 2 columns
combinedata <- cbind(featuressubset,subjectbyactivity)  ##10299 rows, 68 columns
```
7. then get the mean values for the variables by subject and activity
```{r `warning = FALSE` }
by1 <- combinedata$subject
by2 <- combinedata$activity
tidydata <- aggregate(x=combinedata, by = list(by1, by2), FUN = "mean",drop=TRUE)  ##180 rows,70 cols

```
8. clean up up the headers
```{r}
colnames(tidydata)[1:2] <- c("subject","activity")
colnames(tidydata)[3:68] <- gsub("^t","time",colnames(tidydata)[3:68])
colnames(tidydata)[3:68] <- gsub("^f","frequency",colnames(tidydata)[3:68])
colnames(tidydata)[3:68] <- sub("\\-mean\\(\\)\\-","Mean",colnames(tidydata)[3:68])
colnames(tidydata)[3:68] <- sub("\\-mean\\(\\)","Mean",colnames(tidydata)[3:68])
colnames(tidydata)[3:68] <- sub("\\-std\\(\\)\\-","Std",colnames(tidydata)[3:68])
colnames(tidydata)[3:68] <- sub("\\-std\\(\\)","Std",colnames(tidydata)[3:68])
tidydata<-tidydata[1:68] ##180 rows,68 cols
```
9. write out the tidy data
```{r}
setwd(old.dir)
write.table(tidydata, file='./data/tidydata.txt')
```

##Description of the variables in the tiny_data.txt file
The resultant tidydatafinal dataset:
 - Dimensions: 180 rows by 68 cols 
 - It contains the average of each variable for each activity and each subject.
 - Variables names are formatted in camelCase.
 - Examples of renamed variables
 
   "V1"                 became "activities" 
   "tBodyAcc-mean()-X"  became "timeBodyAccMeanX"
   "fBodyAcc-std()-Y"   became "frequencyBodyAccStdY"

### Renamed Variables
- subject
- activity
- timeBodyAccMeanX
- timeBodyAccMeanY
- timeBodyAccMeanZ
- timeBodyAccStdX
- timeBodyAccStdY
- timeBodyAccStdZ
- timeGravityAccMeanX
- timeGravityAccMeanY
- timeGravityAccMeanZ
- timeGravityAccStdX
- timeGravityAccStdY
- timeGravityAccStdZ
- timeBodyAccJerkMeanX
- timeBodyAccJerkMeanY
- timeBodyAccJerkMeanZ
- timeBodyAccJerkStdX
- timeBodyAccJerkStdY
- timeBodyAccJerkStdZ
- timeBodyGyroMeanX
- timeBodyGyroMeanY
- timeBodyGyroMeanZ
- timeBodyGyroStdX
- timeBodyGyroStdY
- timeBodyGyroStdZ
- timeBodyGyroJerkMeanX
- timeBodyGyroJerkMeanY
- timeBodyGyroJerkMeanZ
- timeBodyGyroJerkStdX
- timeBodyGyroJerkStdY
- timeBodyGyroJerkStdZ
- timeBodyAccMagMean
- timeBodyAccMagStd
- timeGravityAccMagMean
- timeGravityAccMagStd
- timeBodyAccJerkMagMean
- timeBodyAccJerkMagStd
- timeBodyGyroMagMean
- timeBodyGyroMagStd
- timeBodyGyroJerkMagMean
- timeBodyGyroJerkMagStd
- frequencyBodyAccMeanX
- frequencyBodyAccMeanY
- frequencyBodyAccMeanZ
- frequencyBodyAccStdX
- frequencyBodyAccStdY
- frequencyBodyAccStdZ
- frequencyBodyAccJerkMeanX
- frequencyBodyAccJerkMeanY
- frequencyBodyAccJerkMeanZ
- frequencyBodyAccJerkStdX
- frequencyBodyAccJerkStdY
- frequencyBodyAccJerkStdZ
- frequencyBodyGyroMeanX
- frequencyBodyGyroMeanY
- frequencyBodyGyroMeanZ
- frequencyBodyGyroStdX
- frequencyBodyGyroStdY
- frequencyBodyGyroStdZ
- frequencyBodyAccMagMean
- frequencyBodyAccMagStd
- frequencyBodyBodyAccJerkMagMean
- frequencyBodyBodyAccJerkMagStd
- frequencyBodyBodyGyroMagMean
- frequencyBodyBodyGyroMagStd
- frequencyBodyBodyGyroJerkMagMean
- frequencyBodyBodyGyroJerkMagStd

A full description of the above data can be found here: 
http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones

##Sources
Google and the Class Forums were of great assistance.

